---
title: "celltype_specific_effect"
author: "Lifan Liang"
date: "2026-04-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Given that there is donor random effect shared across all cell types, the eQTL only exists in one cell type in ground truth. This section validates the ability of different eQTL mapping approaches to detect this cell type specific eQTL for one gene-SNP pair. Simply, the dataset is generated from the follow linear mixed model:

$$
Y_{ijk} = \mu + G_i\beta_j + u_i + \epsilon_{ijk}
$$

Where

* $Y_{ijk}$ is the gene expression for $k$th cell in the $j$th cell type and $i$th donor
* $\mu$ is the baseline gene expression
* $\beta_j$ is the genetic effect in the $j$th cell type, $beta_j$ is nonzero only the first cell type
* $G_i$ is the genotype for the $i$th donor
* $u_i$ is the $i$th donor effect
* $\epsilon$ is the cell-level residual noise

## Balanced cell counts with no cell-level covariates

In this setting, we have 100 donors, 10 cell types, and 100 cells per donor per cell type. PVE of eQTL was 2.5%.


```{r,echo=F}
library(dplyr)
library(lme4)

set.seed(42)
n_donors <- 100
n_celltypes <- 10
n_cells_per_donor_ct <- 100
target_celltype <- "CellType_1"
baseline_expr <- 10.0

# Variance parameters
sigma_donor <- 1.5    # Std dev of the shared donor random effect
sigma_noise <- 0.8    # Std dev of the cell-level residual noise
maf <- 0.3            # Minor Allele Frequency

target_variance_explained <- 0.02 # 5% PVE

var_genotype <- 2 * maf * (1 - maf)
var_other <- (sigma_donor^2) + (sigma_noise^2)

# Solve for Beta (effect size)
var_eqtl <- (target_variance_explained / (1 - target_variance_explained)) * var_other
eqtl_effect_size <- sqrt(var_eqtl / var_genotype)


sim.dat <- function(n_donor, n_celltypes, n_cells_per_donor_ct, maf, sigma_donor, sigma_noise, eqtl_effect_size) {
  donor_ids <- paste0("Donor_", sprintf("%03d", 1:n_donors))
  celltype_ids <- paste0("CellType_", 1:n_celltypes)
  
  genotypes <- rbinom(n_donors, size = 2, prob = maf)
  donor_effects <- rnorm(n_donors, mean = 0, sd = sigma_donor)
  
  df_donors <- data.frame(
    donor_id = donor_ids,
    genotype = genotypes,
    donor_effect = donor_effects
  )
  
  df_cells <- expand.grid(
    cell_index = 1:n_cells_per_donor_ct,
    celltype_id = celltype_ids,
    donor_id = donor_ids,
    stringsAsFactors = FALSE
  ) %>%
    left_join(df_donors, by = "donor_id") %>%
    mutate(
      cell_id = paste(donor_id, celltype_id, cell_index, sep = "_"),
      
      # Apply the mathematically derived effect size ONLY to target cell type
      eqtl_contribution = ifelse(
        celltype_id == target_celltype, 
        eqtl_effect_size * genotype, 
        0
      ),
      
      noise = rnorm(n(), mean = 0, sd = sigma_noise),
      expression = baseline_expr + donor_effect + eqtl_contribution + noise
    ) %>%
    select(cell_id, donor_id, celltype_id, genotype, expression)
}

df_cells <- sim.dat(n_donor, n_celltypes, n_cells_per_donor_ct, maf, sigma_donor, sigma_noise, eqtl_effect_size)

#target_data <- df_cells %>% filter(celltype_id == "CellType_1")
#lmm <- lmer(expression ~ genotype + (1 | donor_id), data = target_data)
#r.squaredGLMM(lmm)
```

### Pseudobulk fixed effect 

We run eQTL mapping on the pseudobulk sample (average gene expression) for each cell type seperately:

$$
\bar Y_{ij} \sim \mu_j + G_i \beta_j + \epsilon_i
$$

```{r}
### Pseudo bulk eQTL mapping
df_pseudobulk <- df_cells %>%
  group_by(donor_id, celltype_id) %>%
  summarize(
    mean_expression = mean(expression),
    
    # We also want to keep the genotype! 
    # Since genotype is identical for all cells from the same donor, 
    # we can just grab the first value.
    genotype = first(genotype),
    
    # Optional: count how many cells went into this average
    n_cells = n(),
    
    .groups = "drop" # Drops the grouping structure afterward
  )

sumstats <- list()
for(i in 1:n_celltypes) {
  n <- paste0("CellType_",i)
  temp <- df_pseudobulk[df_pseudobulk$celltype_id==n,]
  sumstats[[n]] <- summary(lm(mean_expression~genotype, temp))$coefficient[2,]
}

sumstats.ps <- do.call(rbind, sumstats)
sumstats.ps
```

### Joint LMM on pseudobulk samples

Running eQTL mapping across all cell types together with linear mixed model (LMM) to estimate genetic effects:

$$
\bar Y_{ij} \sim \mu_j + \sum X_{j}G_i\beta_j + u_i + \epsilon_i
$$

where $X_j$ is the binary indicator of the $j$th cell type.

```{r, echo=F}
### LMM an the pseudo-bulk sample
sumstats.lmm.ps <- summary(lmer(mean_expression~celltype_id:genotype + (1|donor_id), df_pseudobulk))$coefficients
sumstats.lmm.ps[-1,]
```


### Joint LMM on single cell level

We fit the model below directly using the cell-level simulation data. And we did not observe power increase.

$$
Y_{ijk} \sim \mu +\sum X_{j}G_i\beta_j + u_i + \epsilon_{ijk}
$$

```{r, echo=F}
sumstats.lmm.cell <- summary(lmer(expression~ genotype+celltype_id:genotype + (1|donor_id), df_cells))$coefficients
sumstats.lmm.cell[-1,]
```


The statistical power of detecting the genetic effect in the first cell type is basically the same across all the three models above. It is probably because calculating the standard error is quite similar between the two approaches.

For the pseudobulk fixed effect approach, residual noise ($\epsilon$) has effectively absorbed the random donor effects. Hence $Var(\hat\beta)=\dfrac{Var(G)}{Var(\epsilon)+Var(u_i)}$.

For the LMM approach, the variance of $\beta_j$ is derived as

$$
Var(\hat\beta_j) = (G'V^{-1}G)^{-1}
$$
$$
V = ZGZ' + Var(\epsilon)I= Var(u_i)+Var(\epsilon)
$$

Therefore, $Var(\hat\beta_j)$ is also $Var(u_i)+Var(\epsilon)$, same as that in the fixed effect model.

However, if we included a main genetic effect in the model, then $\beta_{ij}$ in each cell type represents deviation from the $j$th cell type.

### Two-step framework (diffQTL)


```{r,echo=F}
sumstats.dqtl <- matrix(nrow=n_celltypes, ncol=4)
for(j in 1:n_celltypes) {
  dat.diff <- as.data.frame(matrix(nrow=n_donors, ncol=3))
  colnames(dat.diff) <- c("genotype","theta", "theta.se")
  for(i in 1:n_donors) {
    donor <- paste0("Donor_", sprintf("%03.0f",i))
    celltype <- paste0("CellType_", j)
    temp <- df_cells[df_cells$donor_id==donor,]
    temp$target <- 0
    temp$target[temp$celltype_id==celltype] <- 1
    dat.diff[i,2:3] <- summary(lm(expression~target, temp))$coefficients[2,1:2]
    dat.diff$genotype[i] <- temp$genotype[1]
  }
  sumstats.dqtl[j,] <- summary(lm(theta~genotype,dat.diff, weights = 1/dat.diff$theta.se^2))$coefficients[2,]
}
```


## Unbalanced cell count and cell level covariates

The number of cells per donor per cell type varies in the range of $[50,500]$.

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)

# ---------------------------------------------------------
# 1. Define Base Simulation Parameters
# ---------------------------------------------------------
set.seed(42) 
n_donors <- 100
n_celltypes <- 10
target_celltype <- "CellType_1"
baseline_expr <- 10.0

# Variance parameters
sigma_donor <- 1.5    # Shared donor random effect
sigma_noise <- 0.8    # Cell-level residual noise
maf <- 0.3            # Minor Allele Frequency
beta_mito <- -0.1     # Linear effect of mito % on expression (e.g., higher mito = lower expr)

# New Target PVE
target_pve <- 0.02    # 2% variance explained by eQTL

# ---------------------------------------------------------
# 2. Simulate Donor-Level Data
# ---------------------------------------------------------
donor_ids <- paste0("Donor_", sprintf("%03d", 1:n_donors))
celltype_ids <- paste0("CellType_", 1:n_celltypes)

df_donors <- data.frame(
  donor_id = donor_ids,
  genotype = rbinom(n_donors, size = 2, prob = maf),
  donor_effect = rnorm(n_donors, mean = 0, sd = sigma_donor)
)

# ---------------------------------------------------------
# 3. Simulate Unbalanced Cell Counts & Grid
# ---------------------------------------------------------
# Create the base grid of all donors and cell types
df_grid <- expand.grid(
  donor_id = donor_ids, 
  celltype_id = celltype_ids, 
  stringsAsFactors = FALSE
)

# Randomly sample the number of cells (50 to 500) for each donor-celltype pair
df_grid$n_cells <- sample(50:500, size = nrow(df_grid), replace = TRUE)

# Expand the grid into single cells
df_cells <- df_grid %>%
  uncount(n_cells) %>%                           # Duplicates rows based on n_cells
  group_by(donor_id, celltype_id) %>%
  mutate(cell_index = row_number()) %>%          # Assign unique index within the cluster
  ungroup() %>%
  left_join(df_donors, by = "donor_id") %>%
  mutate(cell_id = paste(donor_id, celltype_id, cell_index, sep = "_"))

# ---------------------------------------------------------
# 4. Simulate Mitochondria Covariate (Mixture Distribution)
# ---------------------------------------------------------
n_total_cells <- nrow(df_cells)

# Create a boolean mask where 90% of cells are TRUE (low mito)
is_low_mito <- runif(n_total_cells) < 0.90

# Generate mito percentages: 
# 90% drawn uniformly from 0 to 5%
# 10% drawn uniformly from 5% to 25%
df_cells$mito_percent <- ifelse(
  is_low_mito,
  runif(n_total_cells, min = 0, max = 5),
  runif(n_total_cells, min = 5, max = 25)
)

# Calculate the strict linear effect of mito on expression
df_cells$mito_effect <- df_cells$mito_percent * beta_mito

# ---------------------------------------------------------
# 5. Dynamically Calculate eQTL Effect Size
# ---------------------------------------------------------
# To hit exactly 2% PVE, we must account for the variance added by the mito effect
# We calculate the empirical variance of the mito effect specifically in the target cell type
target_mito_variance <- var(df_cells$mito_effect[df_cells$celltype_id == target_celltype])

var_genotype <- 2 * maf * (1 - maf)
var_other <- (sigma_donor^2) + (sigma_noise^2) + target_mito_variance

# Solve for Beta
var_eqtl <- (target_pve / (1 - target_pve)) * var_other
eqtl_effect_size <- sqrt(var_eqtl / var_genotype)

cat(sprintf("Calculated eQTL effect size (Beta) to explain %.0f%% variance: %.4f\n", 
            target_pve * 100, eqtl_effect_size))

# ---------------------------------------------------------
# 6. Finalize Gene Expression
# ---------------------------------------------------------
df_cells <- df_cells %>%
  mutate(
    eqtl_contribution = ifelse(
      celltype_id == target_celltype, 
      eqtl_effect_size * genotype, 
      0
    ),
    noise = rnorm(n(), mean = 0, sd = sigma_noise),
    expression = baseline_expr + donor_effect + mito_effect + eqtl_contribution + noise
  ) %>%
  select(cell_id, donor_id, celltype_id, genotype, mito_percent, expression)

# ---------------------------------------------------------
# 7. Validation Check
# ---------------------------------------------------------
# Check the mito percentage distribution
cat(sprintf("\nProportion of cells with < 5%% mitochondria: %.1f%%\n", 
            mean(df_cells$mito_percent < 5) * 100))

# Check cell count ranges
cat(sprintf("Min cells per group: %d | Max cells per group: %d\n", 
            min(table(df_cells$donor_id, df_cells$celltype_id)), 
            max(table(df_cells$donor_id, df_cells$celltype_id))))
```

```{r, echo=F}
sumstats.lmm.cell <- summary(lmer(expression~ celltype_id:genotype + (1|donor_id), df_cells))$coefficients
sumstats.lmm.cell[-1,]
```
