---
title: "context marginal eQTL association"
author: "Lifan Liang"
date: "2026-04-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Simulation

We directly simulated the pseudobulk samples where only one out of 10 contexts has nonzero genetic effect:

$$
Y_{ij} = \mu + 0.5 \cdot X_1 G_i + \sum_{j=2}^{10}0\cdot X_jG_i + u_i + \epsilon_{ij}
$$

Where

* $Y_{ij}$ is the averaged gene expression for the $j$th context and $i$th donor
* $\mu$ is the baseline gene expression
* $G_i$ is the genotype for the $i$th donor
* $X_j$ is the binary indicator for the $j$th cell type, $X_j=1$ when $Y_{ij}$ is in the same cell type
* $u_i$ is the $i$th donor effect
* $\epsilon$ is the residual noise for the $j$th context and $i$th donor


```{r}
## Direct generation of pseudobulk
library(dplyr)
library(lme4)

set.seed(45)
n_donors <- 500
n_celltypes <- 10
target_celltype <- "CellType_1"
baseline_expr <- 10.0

# Variance parameters
sigma_donor <- 0.8    # Std dev of the shared donor random effect
sigma_noise <- 0.8    # Std dev of the cell-level residual noise
maf <- 0.3            # Minor Allele Frequency

target_variance_explained <- 0.05 # 5% PVE

var_genotype <- 2 * maf * (1 - maf)
var_other <- (sigma_donor^2) + (sigma_noise^2)

# Solve for Beta (effect size)
var_eqtl <- (target_variance_explained / (1 - target_variance_explained)) * var_other
eqtl_effect_size <- sqrt(var_eqtl / var_genotype)

#eqtl_effect_size <- 0.5
sim.dat <- function(n_donor, n_celltypes, n_cells_per_donor_ct, maf, sigma_donor, sigma_noise, eqtl_effect_size) {
  donor_ids <- paste0("Donor_", sprintf("%03d", 1:n_donors))
  celltype_ids <- paste0("CellType_", 1:n_celltypes)
  
  genotypes <- rbinom(n_donors, size = 2, prob = maf)
  donor_effects <- rnorm(n_donors, mean = 0, sd = sigma_donor)
  
  df_donors <- data.frame(
    donor_id = donor_ids,
    genotype = genotypes,
    donor_effect = donor_effects
  )
  
  df_bulk <- expand.grid(
    celltype_id = celltype_ids,
    donor_id = donor_ids,
    stringsAsFactors = FALSE
  ) %>%
    left_join(df_donors, by = "donor_id") %>%
    mutate(
      # Apply the mathematically derived effect size ONLY to target cell type
      eqtl_contribution = ifelse(
        celltype_id == target_celltype, 
        eqtl_effect_size * genotype, 
        0
      ),
      
      noise = rnorm(n(), mean = 0, sd = sigma_noise),
      expression = baseline_expr + donor_effect + eqtl_contribution + noise
    ) %>%
    select(donor_id, celltype_id, genotype, expression, donor_effect)
}

df_ps <- sim.dat(n_donor, n_celltypes, n_cells_per_donor_ct, maf, sigma_donor, sigma_noise, eqtl_effect_size)
```


## Pseudobulk eQTL mapping

```{r}
n_iter <- 100
ps_eqtl_df <- matrix(nrow=n_iter, ncol=n_celltypes)
for(iter in 1:n_iter) {
  df_ps <- sim.dat(n_donor, n_celltypes, n_cells_per_donor_ct, maf, sigma_donor, sigma_noise, eqtl_effect_size)
  sumstats <- list()
  for(i in 1:n_celltypes) {
    n <- paste0("CellType_",i)
    temp <- df_ps[df_ps$celltype_id==n,]
    sumstats[[n]] <- summary(lm(expression~genotype, temp))$coefficient[2,]
  }
  
  sumstats.ps <- do.call(rbind, sumstats)
  ps_eqtl_df[iter,] <- sumstats.ps[,1]
}
boxplot(ps_eqtl_df, main="Effect size distribution for 10 contexts")
lines(c(0.2,1.7), rep(eqtl_effect_size,2), col="red", lwd=2)
lines(c(1.5,12), rep(0,2), col="blue", lwd=2)
```

## LMM eQTL mapping

```{r}
set.seed(42)
n_iter <- 100
lmm_eqtl_df <- matrix(nrow=n_iter, ncol=3)
colnames(lmm_eqtl_df) <- c("no_main_eff", "besides_main_eff", "plus_main_eff")
for(iter in 1:n_iter) {
  df_ps <- sim.dat(n_donor, n_celltypes, n_cells_per_donor_ct, maf, sigma_donor, sigma_noise, eqtl_effect_size)
  df_ps$target_geno <- 0
  df_ps$target_geno[df_ps$celltype_id=="CellType_1"] <- 1
  temp <- summary(lmer(expression~genotype+target_geno+(1|donor_id), df_ps))$coefficient
  lmm_eqtl_df[iter,2] <- temp[3,1]
  lmm_eqtl_df[iter,1] <- summary(lmer(expression~target_geno+(1|donor_id), df_ps))$coefficient[2,1]
  lmm_eqtl_df[iter,3] <- temp[3,1] + temp[2,1]
}
boxplot(lmm_eqtl_df)
```

