---
title: "Updated real data analysis with S-MIRAGE"
author: "Lifan Liang"
date: "2026-03-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
<style>
.scroll-300 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
</style>

## Previous issues

We applied S-MIRAGE to REGENERON EWAS for the trait of LDL-C. The [initial results](MIRAGE_extension.html#Application_to_REGENIE_dataset) discovered 9 genes, whereas the burden test from the original study found 14 genes. 

One major issue is the winners' curse for estimating $\mu$ and $\sigma$. These parameters will be severely overestimated using a truncated set. This is obvious with simulation experiments below. We synthesized 10,000 variants with true genetic effects and filtered the variants with Bonferroni cutoff (P<5e-7). Then we estimated $\sigma$ and $\mu$ from the filtered summary statistics. The estimated parameters have much larger magnitude than the ground truth.


```{r, class.source="scroll-300"}
# 1. Simulate a full genome of 10,000 variants
set.seed(42)
total_M <- 10000
true_mu <- -0.05
true_sigma2 <- 0.04

# True biological effects
true_beta <- rnorm(total_M, mean = true_mu, sd = sqrt(true_sigma2))

# Add measurement noise (standard errors around 0.1)
se_full <- runif(total_M, 0.08, 0.12)
beta_hat_full <- true_beta + rnorm(total_M, mean = 0, sd = se_full)

# 2. Apply a strict p-value filter (e.g., p < 0.01)
z_scores <- beta_hat_full / se_full
p_values <- 2 * pnorm(-abs(z_scores))

# Keep only the significant variants
sig_idx <- which(p_values < 5e-7)
obs_beta <- beta_hat_full[sig_idx]
obs_se <- se_full[sig_idx]

sigma2_hat <- var(obs_beta)-mean(obs_se^2)
mu_hat <- weighted.mean(obs_beta, w=1/(sigma2_hat+obs_se))
  
cat("Variants passing filter:", length(obs_beta), "out of", total_M, "\n")
cat("Ground truth variance of genetic effects", sqrt(true_sigma2), "\n")
cat("Naive Variance Estimate (Winner's Curse):", sqrt(sigma2_hat), "\n")
cat("Ground truth genetic effects:",true_mu, "\n")
cat("Naive Mu estimate (Winner's Curse):", mu_hat," \n\n")

```

## Correcting winner's curse with EM algorithm

A more conceptually straightforward approach would be estimating $\mu$ and $\sigma$ inside the EM algorithm where we estimated $\pi$ and $\delta$. However, there is no close-form solution to the MLE of $\mu$ and $\sigma$. Optimization involve frequent computation of the likelihood function over half a million variants. The time complexity of such computation is too high for a single M-step.

It is more feasible to use variants above the Bonferroni threshold as true risk variants. But we need to account for the risk variants filtered by the Bonferroni threshold. EM algorithm can address this problem by treating the filtered causal variants as the missing data. And we are essentially finding $\hat\mu$ and $\hat\sigma$ that maximize the complete data log likelihood:

$$
\begin{align}
\mathcal{ll}_{complete}(\mu,\sigma) &= \sum_{i=1}^N \mathcal{ll}(\hat\beta_i;\mu,\sigma,s_i) + \sum_{j=1}^{M-N} \mathcal{ll}(\hat\beta_j;\mu,\sigma,s_j)
\end{align}
$$

where $N$ is the number of true risk variants that pass the filter, and $M$ is the total number of true risk variants. In the M-step, we simply used `optim()` to find the $\mu$ and $\sigma$ that maximzie the complete data likelihood. The real problem is how to impute $\hat\beta$ for the missing data given $\mu$ and $\sigma$ in the E-step. We can expand the log likelihood function related to $\hat\beta$:

$$
\begin{align}
\mathcal{ll}_{missing} &\propto \sum_{j=1}^{M-N}\dfrac{(\hat\beta_j - \mu)^2}{-2(\sigma^2+s_j^2)}\\
&\propto \sum_{j=1}^{M-N}(\hat\beta_j^2 - 2\mu \hat\beta_j) \\
&= (M-N)\left(E[\hat\beta^2] - 2 \mu E[\hat\beta]\right)
\end{align}
$$

To compute the complete data likelihood, we only need the first moment and the second moment of $\hat\beta$ conditioned on failing the filter. A simpler and more general approach would be to perform Monte Carlo sampling to estimate the two moments. But that would be computationally expensive when M grows large. Fortunately, we can derive the analytical solution from the [statistical results for the truncated Normal distribution](10.1080/00031305.1999.10474490):

$$
E[\hat\beta_j] = \mu + \sqrt{\sigma^2 + s_j^2} \cdot \left( \dfrac{\phi(\alpha_j)-\phi(\gamma_j)}{T_j} \right)
$$
where $\phi$ is PDF for standard Normal, $\alpha$ and $\gamma$ are the truncation boundary after Z transformation (e.g. $\alpha_j=\dfrac{\hat\beta_j - \mu}{\sqrt{\sigma^2 + s_j^2}}$), and $T_j$ is the proportion of $Normal(\mu, \sigma^2+s^2)$ that falls within the Null truncation ($\Phi(\alpha_j)-\Phi(\gamma_j)$).

$$
E[\hat\beta_j^2] = Var(\hat\beta) + (E[\hat\beta])^2
$$


$$
Var(\hat\beta_j) = V_j\left[ 1+\dfrac{\alpha_j \phi(\alpha_j) - \beta_j\phi({\beta_j})}{T_j} - \left( \dfrac{\phi(\alpha)-\phi(\beta)}{T_j}\right)^2 \right]
$$

```{r, class.source="scroll-300"}
em_winner_curse <- function(beta_hat, se, p_threshold, total_M, 
                            init_mu = 0, init_sigma2 = 0.01, 
                            max_iter = 500, tol = 1e-6) {
  
  # 1. Basic Setup & Validation
  N_obs <- length(beta_hat)
  N_miss <- total_M - N_obs
  
  if (N_miss <= 0) {
    stop("total_M must be strictly greater than the number of observed variants.")
  }
  
  # Calculate the Z-score truncation threshold
  c_thresh <- qnorm(1 - p_threshold / 2)
  
  # Approximate the standard error for the missing variants
  s_miss <- median(se)
  
  # Initialize parameters
  mu <- init_mu
  sigma2 <- init_sigma2
  
  for (iter in 1:max_iter) {
    mu_old <- mu
    sigma2_old <- sigma2
    
    # ========================================================
    # E-STEP: Impute expected moments for the missing variants
    # ========================================================
    V_miss <- sigma2 + s_miss^2
    sd_miss <- sqrt(V_miss)
    
    # Truncation boundaries
    a <- -c_thresh * s_miss
    b <-  c_thresh * s_miss
    
    # Standardize boundaries based on current parameter estimates
    alpha <- (a - mu) / sd_miss
    beta  <- (b - mu) / sd_miss
    
    # Normalization constant Z (probability of falling in the missing zone)
    Z <- pnorm(beta) - pnorm(alpha)
    
    # Prevent division by zero if Z gets vanishingly small
    if (Z < 1e-15) Z <- 1e-15 
    
    # Calculate E[z] and E[z^2] for the standard normal
    phi_alpha <- dnorm(alpha)
    phi_beta  <- dnorm(beta)
    
    E_z  <- (phi_alpha - phi_beta) / Z
    E_z2 <- 1 + (alpha * phi_alpha - beta * phi_beta) / Z
    
    # Unstandardize to get E[beta] and E[beta^2] for a single missing variant
    E_beta_miss  <- mu + sd_miss * E_z
    E_beta2_miss <- mu^2 + 2 * mu * sd_miss * E_z + V_miss * E_z2
    
    # ========================================================
    # M-STEP: Maximize the expected complete data log-likelihood
    # ========================================================
    # We define the negative Q-function to minimize using optim()
    neg_Q <- function(params) {
      curr_mu <- params[1]
      curr_sigma2 <- params[2]
      
      # 1. Log-likelihood of the OBSERVED data
      V_obs <- curr_sigma2 + se^2
      ll_obs <- -0.5 * sum(log(V_obs) + (beta_hat - curr_mu)^2 / V_obs)
      
      # 2. Expected log-likelihood of the MISSING data
      V_m <- curr_sigma2 + s_miss^2
      
      # E[(beta - mu)^2] = E[beta^2] - 2*mu*E[beta] + mu^2
      exp_sq_diff <- E_beta2_miss - 2 * curr_mu * E_beta_miss + curr_mu^2
      ll_miss <- -0.5 * N_miss * (log(V_m) + exp_sq_diff / V_m)
      
      # Return negative sum for minimization
      return(-(ll_obs + ll_miss))
    }
    
    # Optimize Q with respect to mu and sigma2 (keeping sigma2 > 0)
    opt <- optim(par = c(mu, sigma2), 
                 fn = neg_Q, 
                 method = "L-BFGS-B", 
                 lower = c(-Inf, 1e-8))
    
    mu <- opt$par[1]
    sigma2 <- opt$par[2]
    
    # ========================================================
    # Convergence Check
    # ========================================================
    if (max(abs(mu - mu_old), abs(sigma2 - sigma2_old)) < tol) {
      cat("Converged successfully in", iter, "iterations.\n")
      break
    }
  }
  
  if (iter == max_iter) cat("Warning: Reached max_iter without fully converging.\n")
  
  return(list(
    estimated_mu = mu,
    estimated_sigma2 = sigma2,
    estimated_sigma = sqrt(sigma2),
    iterations = iter
  ))
}


```

We applied this EM algorithm to the data. The estimate for $\mu$ and $\sigma$ are much closer to ground truth than the naive estimate.

```{r}
results <- em_winner_curse(
  beta_hat = obs_beta, 
  se = obs_se, 
  p_threshold = 5e-7, 
  total_M = total_M,
  init_mu = mean(obs_beta),     # Good starting point
  init_sigma2 = var(obs_beta)   # Good starting point
)

cat("EM estimate for genetic effect:", results$estimated_mu, "\n")
cat("EM estimate for variance of genetic effect:", results$estimated_sigma2, "\n\n")
```


```{r}
barplot(c(Naive=sigma2_hat, EM=results$estimated_sigma2), main="Estimated sigma2 in the winners' curse")
lines(c(0,10),c(0.04,0.04), col="red", , lty=2)
text(0.5,0.05,"Ground truth", col="red")
```

```{r}
barplot(c(Naive=mu_hat, EM=results$estimated_mu), main="Estimated mu in the winners' curse")
lines(c(0,10),c(-0.1,-0.1), col="red", lty=2)
text(0.5, -0.09,"Ground truth", col="red")
```


## Updated with AlphaMissense annotation

We currently have 78235 LoF variants and 328440 missense variants. The missense variants may need more detail annotations. So we use the prediction from AlphaMissense to annotate the missense variants. The missense variants were further divided into three categories: (1) likely pathogenic; (2) ambiguous; (3) likely benign.


```{r}
library(ggplot2)
df <- data.frame(category=c("LoF","Pathogenic","Ambiguous","Benigh"), 
                 mean=-0.06132739, -0.04140436, 0.01981949, -0.01473132,
                 SE=0.3368726, 0.2492694, 0.1797131, 0.2009707)

ggplot(df, aes(x = category, y = mean)) +
  geom_point(size = 3, color = "blue") +
  geom_errorbar(aes(ymin = mean - SE, ymax = mean + SE), width = 0.1) +
  theme_classic()
```

Using this set of parameters to estimate the risk proportion of genes and variants, we have roughly 1.77% of genes being risk genes. The risk proportion for each category of variants are:

```{r}
barplot(c("LoF"=0.28383799, "Pathogenic"=0.19723679, "Ambiguous"=0.06492431, "Benign"=0.07392038),
        las=2, col=1:4)
```

This set of parameters yields 13 risk genes. Missing ZNF318 compared to the previous results.


## Updated with estimating $\sigma$ and $\mu$ in the main EM algorithm

We applied a more loose threshold, P < 1e-5, to extract around 30 variants for the category of LoF and missense, respectively. Then we only used the top variant per gene. 7 variants remains for LoF and 10 variants for missense.

Use the most severe consequence of the variants and only use LoF and missense. Mu and Sigma were estimated for the two categories separately.

$$
\begin{align}
\sigma_{LoF} = 0.5, && \sigma_{missense}=0.2 \\
\mu_{LoF} = -0.1, && \mu_{missense} = 0
\end{align}
$$

We compared our methods with ACAT and the sumstats version of burden test.

```{r}
library(VennDiagram)

tri.res <- readRDS("data/three_type_LDL_risk_genes.RDS")
p1 <- venn.diagram(tri.res, filename = NULL)
grid.newpage()
grid.draw(p1)
```

Burden test has found two more genes. Among the six genes uniquely discovered by the burden test, HMGCR is known to be related to the transportation of LDL. This gene is missed by S-MIRAGE because several variants have large positive Z score. But S-MIRAGE has estimated a negative $\mu$ for the variant effects.

One single Normal distribution assumes that variant effects are mostly small. It does not hold for rare variants that convey risks, especially the LoF variants. Moreover, although most variants' impact on genes are mostly negative, the gene's association to the trait could be positive or negative. This would flip the signs of the variant effects on the trait. Therefore, it would be more reasonable to adopt the symmetric bimodal distribution to model the genetic effects of the risk variants.

## Symmetric bimodal distribution for risk variants

When the variant is a risk variant, $\hat\beta$ is sampled from the following procedure:

$$
P(\beta_i|\mu,\sigma) = 0.5 \mathcal N(\beta_i; \mu,\sigma^2) + 0.5\mathcal N(\beta_i; -\mu, \sigma^2)
$$

$$
P(\hat\beta_i|\beta_i, s_i) = \mathcal N(\hat\beta_i;\beta_i,s_i^2)
$$

### EM algorithm for inference

In the E step, we compute the posterior of the variants belonging to one of the modes if this is a risk variant ($\gamma_i$), the posterior mean and variance of variant effect ($\bar\beta_i$ and $Var(\bar\beta_i)$) given that the variant is a risk variant, the variance of the posterior of risk variants ($z_i$), and the posterior of risk genes ($u_j$).

$$
\begin{align}
&L_{1+} = \mathcal N(\hat\beta_i;\mu,\sigma^2+s_i^2) \\
&L_{1-} = \mathcal N(\hat\beta_i;-\mu,\sigma^2+s_i^2) \\
&\gamma_{i+} = \dfrac{L_{1+}}{L_{1+}+L_{1-}} \\
&w_i = \dfrac{\sigma^2}{\sigma^2+s_i^2} \\
&\mu_{i+} = (1-w_i)\mu + w_i \hat\beta \\
&\bar\beta_i = \gamma_{i-}\cdot\mu_{i-} + \gamma_{i+}\cdot\mu_{i+} \\
&\Sigma_i = \dfrac{\sigma^2 s_i^2}{\sigma^2+s_i^2} \\
&Var(\bar\beta_i) = \Sigma_i + \gamma_{i+}(\mu_{i+}-\bar\beta_i)^2 +\gamma_{i-}(\mu_{i-}-\bar\beta_i)^2 \\
&B_i = \dfrac{0.5\pi(L_{1+} + L_{1-})}{0.5\pi(L_{1+} + L_{1-}) + (1-\pi)L_0}
\end{align}
$$

The rest is the same as the previous model. We introduced extra steps to compute the posterior of genetic effects under the alternative model because we can use them to update $\mu$ and $\sigma$ for each variant category with a close-form solution in the M-step. When the $\bar\beta$ is fixed, we can directly estimate $\mu$ and $\sigma$ with the close-form MLE over the complete data likelihood, ignoring $s_i$.

$$
\mu = \dfrac{\sum_i z_i ( \gamma_{i+}\cdot\mu_{i+} - \gamma_{i-}\cdot\mu_{i-})}{\sum_i z_i}
$$

$$
\sigma = \dfrac{\sum_i z_i Var(\bar\beta_i)}{\sum_i z_i}
$$




## Application to neutrophil counts

We applied the same procedure to the summary statistics of the trait of neutrophil counts. The estimate of genetic parameter from 10 risk variants above the Bonferroni threshold were:

```{r,echo=F}
params <- readRDS("data/neutrophil_params.RDS")
```


$$
\begin{align}
\mu=0.01, &&  \sigma=0.11
\end{align}
$$

We estimated the proportion of risk genes and risk variants with the genetic effect parameter. Risk variants from all categories converged to around 27%. The proportion of risk genes converged to 5.5%.

```{r,echo=F}
matplot(params$pi.trac, xlab="EM iteration", 
        ylab="Proportion of risk variants for each category")
```


With the parameters above, we inferred the posterior of risk genes for all the 17723 genes. 27 of them has PIP>80%. Compared to the 16 risk genes identified in the original study, we have 7 genes in common.

```{r,echo=F}
post.gene <- readRDS("data/neutrophil_pip.RDS")
risk.gene <- names(post.gene)[post.gene>0.8]
print(post.gene[post.gene>0.8])
```

```{r,echo=F}
library(VennDiagram)

orig <- c("GNB1","CSF3R","CXCR4","CXCR2","CXCL6","CXCL2","TET2","BRAF","JAK2","SH2B3","NF1","MPO","SLC9A3R1","S1PR4","IL17RA","CHEK2")

p1 <- venn.diagram(list(REGENIE=orig, MIRAGE=risk.gene),filename=NULL)
grid.newpage()
grid.draw(p1)
```




